---
title: "Stawberries 3"
author: MA615
date: 2024 Sept 30
format:
  html:
    embed-resources: true
editor: visual
---

# Version 5

We ditch the counties

# Preparing data for analysis

Acquire, explore, clean & structure, EDA

### Data cleaning and organization

["An introduction to data cleaning with R" by Edwin de Jonge and Mark van der Loo](https://cran.r-project.org/doc/contrib/de_Jonge+van_der_Loo-Introduction_to_data_cleaning_with_R.pdf)

["Problems, Methods, and Challenges in Comprehensive Data Cleansing" by Heiko Müller and Johann-Christoph Freytag](https://www.researchgate.net/profile/Heiko-Mueller/publication/228929938_Problems_methods_and_challenges_in_comprehensive_data_cleansing/links/09e415101b58541e2c000000/Problems-methods-and-challenges-in-comprehensive-data-cleansing.pdf)

## Strawberries

**Questions**

-   Where they are grown? By whom?

-   Are they really loaded with carcinogenic poisons?

-   Are they really good for your health? Bad for your health?

-   Are organic strawberries carriers of deadly diseases?

-   When I go to the market should I buy conventional or organic strawberries?

-   Do Strawberry farmers make money?

-   How do the strawberries I buy get to my market?

## The data

The data set for this assignment has been selected from:

\[[USDA_NASS_strawb_2024SEP25](https://quickstats.nass.usda.gov) The data have been stored on NASS here: [USDA_NASS_strawb_2024SEP25](https://quickstats.nass.usda.gov/results/28141CE5-7429-36C3-A385-33ACC8030763)

and has been stored on the blackboard as strawberries25_v3.csv.

## read and explore the data

Set-up

```{r}
#| label: load libraries and set options
#| warning: false
#| message: false
#| echo: false

library(knitr)  
library(kableExtra)
library(tidyverse)


```

Read the data and take a first look

```{r}
#| label: read data - glimpse 

strawberry <- read_csv("strawberries25_v3.csv", col_names = TRUE, show_col_types = FALSE )

## glimpse(strawberry)
```

Ditch the counties

```{r}
#| label: ditch the counties

unique(strawberry$`Geo Level`)

strawberry <- strawberry |> 
  filter(`Geo Level`== "NATIONAL" | `Geo Level`== "STATE")
  

```

I have 5359 rows and 21 columns.

## remove columns with a single value in all rows

```{r}
#|label: function def - drop 1-item columns

drop_one_value_col <- function(df){ 
  df_id <- ensym(df)
  msg = paste("Looking for single value columns in data frame: ",as.character(df_id) )
  print(msg)
  ## takes whole dataframe
drop <- NULL
val <- NULL
## test each column for a single value
for(i in 1:dim(df)[2]){   
  if(distinct(df[,i]) |> count() == 1){
    drop <- c(drop, i)
    val <- c(val, df[1,i])
  }
} 

if(is.null(drop)){
  print("No columns dropped")
  return(df)}else{
   print("Columns dropped:")
   # print(colnames(df)[drop])
   print(unlist(val))
   df <- df[, -1*drop]
   return(df)
   }
}


## use the function

strawberry <- strawberry |> drop_one_value_col()


```

Split the census data from the survey data. drop single value columns

## Census data first

```{r}
#| label: srawberries split census, survey


straw_cen <- strawberry |> filter(Program=="CENSUS")

straw_sur <- strawberry |> filter(Program=="SURVEY")

straw_cen <- straw_cen |> drop_one_value_col()

straw_sur <- straw_sur |> drop_one_value_col()


```

```{r}
#| label: straw_cen split cols

straw_cen <- straw_cen |>
  separate_wider_delim(  cols = `Data Item`,
                         delim = " - ",
                         names = c("strawberries",
                                 "Category"),
                         too_many = "error",
                         too_few = "align_start"
                       )

```

```{r}
#| label: isolate organic

# 
# unique(straw_cen$strawberries)
# straw_cen$strawberries |> str_which("STRAWBERRIES") |> length()
# straw_cen$strawberries |> str_which("STRAWBERRIES, ORGANIC") |> length()
# straw_cen$strawberries |> str_which("STRAWBERRIES, ORGANIC, FRESH MARKET") |> length()
# straw_cen$strawberries |> str_which("STRAWBERRIES, ORGANIC, PROCESSING" ) |> length()
# 
# ## count the cases
# straw_cen$strawberries |> str_which("ORGANIC") |> length()
# straw_cen$strawberries |> str_which("FRESH MARKET") |> length()
# straw_cen$strawberries |> str_which("PROCESSING") |> length()

straw_cen <- straw_cen |>
  separate_wider_delim(  cols = strawberries,
                         delim = ", ",
                         names = c("strawberries",
                                 "ORGANIC",
                                 "organic_detail"),
                         
                         too_many = "error",
                         too_few = "align_start"
                       )


straw_cen <- straw_cen |> drop_one_value_col()

## how many organic rows?

organic_cen <- straw_cen |> filter(ORGANIC == "ORGANIC")

sum(is.na(straw_cen$ORGANIC))

straw_cen <- straw_cen[(is.na(straw_cen$ORGANIC)),]


straw_cen <- straw_cen |> drop_one_value_col()



```

Note that straw_cen has only one year: 2022

Current stats Census date has been isolated and split between Organic and Conventional strawberries

############################################ 

```{r}
#| label: explore straw_cen$Category

straw_cen <- straw_cen |>
  separate_wider_delim(  cols = `Category`,
                         delim = " ",
                         names = c("COL1",
                                 "COL2"),
                         too_many = "merge",
                         too_few = "align_start"
                       )


straw_cen$COL2 <- str_replace(straw_cen$COL2,"WITH ","")

straw_cen <- straw_cen |> rename(Measure = COL1, Bearing_type= COL2)

```

```{r}
#| label: explore straw_cen$Domain & Domain Category


## remove AREA GROWN and parens 
## change NOT SPECIFIEC TO TOTAL

straw_cen <- straw_cen |> rename(size_bracket = `Domain Category`)

straw_cen$size_bracket <- str_replace(straw_cen$size_bracket, "NOT SPECIFIED", "TOTAL")

straw_cen$size_bracket <- str_replace(straw_cen$size_bracket, "AREA GROWN: ", "")

```

```{r}
#| label: explore organic_cen

organic_cen <- organic_cen |> drop_one_value_col()



```

Now examine the Survey data

```{r}
#| label: Survey data

## Data Item

## unique(straw_sur$`Data Item`)

straw_sur1 <- straw_sur |>  separate_wider_delim(cols = `Data Item`,
                                                delim = ", ",
                                                names = c("straw",
                                                         "mkt",
                                                         "measure",
                                                         "other"
                                                         ),
                                                too_many = "merge",
                                                too_few = "align_start")
                                                
                                                  
straw_sur2 <- straw_sur1 |> separate_wider_delim(cols = "straw", 
                                                 delim = " - ",
                                                 names = c("straw",
                                                           "more"),
                                                 too_many = "merge",
                                                 too_few = "align_start"
                                                 )
  

rm(straw_sur, straw_sur1)

```

Shift data into alignment

```{r}
#| label: Shift data on a row

## function shift_loc
## Moves adjacent data cells in a data.frame on a single row
## Use this function to fix alignment problems after separating 
## columns containing multiple columns of data. 

## Of course the working assumption is that there is room in the 
## data frame for the data you're shifting.
##
## The data cells that are empty after the data shift are NA.
## 
## Input paramaters
## 
## df -- data frame
## col_name -- name of colume where the left-most data item is located
## dat_name -- name of data item in the column
## num_col -- the number of columns is the same as the number of
##            adjacent data to be moved.
## num_shift -- the number of rows to move the data 
##


shift_loc <- function(df, col_name, dat_name, num_col, num_shift){
 # browser()
  col_num = which(colnames(df) == col_name)
  row_num = which(df[,col_num] == dat_name)  ## calcs a vector of rows
  
  for(k in 1:length(row_num)){
  d = rep(0,num_col) ## storage for items to be moved
  for(i in 1:num_col){
    d[i] = df[row_num[k], col_num + i - 1]
  }
  for(i in 1:num_col){
    ra = row_num[k]
    cb = col_num + i - 1
    df[ra, cb] <-  NA
  }
  for(j in 1:num_col){
    rc = row_num[k]
    cd = col_num + j - 1 + num_shift
    df[rc, cd] = d[j]
  }
  }
 # sprintf("Rows adjusted:")
  # print("%d",row_num)
  return(df)
}




straw_sur2 <- straw_sur2 |> shift_loc("more", "PRICE RECEIVED", 2, 1 )


straw_sur2 <- straw_sur2 |> shift_loc("more", "ACRES HARVESTED", 1, 1 )

straw_sur2 <- straw_sur2 |> shift_loc("more", "ACRES PLANTED", 1, 1 )

straw_sur2 <- straw_sur2 |> shift_loc("more", "PRODUCTION", 2, 1 )

straw_sur2 <- straw_sur2 |> shift_loc("more", "YIELD", 2, 1 )

straw_sur2 <- straw_sur2 |> shift_loc("more", "APPLICATIONS", 3, 1 )

straw_sur2 <- straw_sur2 |> shift_loc("more", "TREATED", 3, 1 )

straw_sur2 <- straw_sur2 |> drop_one_value_col()


```

Examine Domain

```{r}
#| label: use domain to make chem, fert, total data frames



unique(straw_sur2$Domain)

straw_sur2 <- straw_sur2 |>  separate_wider_delim(cols = Domain,
                                                delim = ", ",
                                                names = c("col1",
                                                         "col2"),
                                                too_many = "merge",
                                                too_few = "align_start")


unique(straw_sur2$col1)


survey_d_total <- straw_sur2 |>  filter(col1 == "TOTAL")

survey_d_chem <- straw_sur2 |>  filter(col1 == "CHEMICAL")

survey_d_fert <- straw_sur2 |>  filter(col1 == "FERTILIZER")


```

now look at totals

```{r}


survey_d_total <- survey_d_total |> drop_one_value_col()

### align terms



survey_d_total <- survey_d_total |> shift_loc("measure", "MEASURED IN $ / CWT", 1, 1 )


survey_d_total <- survey_d_total |> shift_loc("measure", "MEASURED IN $", 1, 1 )


survey_d_total <- survey_d_total |> shift_loc("measure", "MEASURED IN CWT", 1, 1 )

survey_d_total <- survey_d_total |> shift_loc("measure", "MEASURED IN TONS", 1, 1 )


survey_d_total <- survey_d_total |> shift_loc("measure", "MEASURED IN CWT / ACRE", 1, 1 )

survey_d_total <- survey_d_total |> shift_loc("measure", "MEASURED IN TONS / ACRE", 1, 1 )


#### split the mkt column


survey_d_total <- survey_d_total |>  
  separate_wider_delim(cols = mkt,
                      delim = " - ",
                      names = c("col3",
                                "col4"),
                   too_many = "merge",
                    too_few = "align_start")






```

there are two markets for Strawberries -- Fresh Marketing and Processing

make a table for each

from the Survey Totals

we have reports for

Markets: Fresh and Processing Operations: Growing and Production

```{r}

survey_d_total <- survey_d_total |> 
  group_by(Year) |> 
  group_by(Period) |>
  group_by(`Geo Level`) |>
  group_by(State) |> 
  group_by(col3)





```
